\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Retrieval-Augmented Generation (RAG) is among the most effective ways to tie language models to scientific evidence. Even so, many open-source RAG implementations still tend to hallucinate and drift out-of-context in specialised domains. Earth Observation (EO) is one such domain: queries typically combine multiple sensors, biophysical variables, and strict spatial and temporal boundaries. Even when the correct documents are retrieved, the generated context often fails to reflect the specific aspects of the question, degrading the final answer.

To address these issues, we propose an ontology-based three-agent pipeline that extends an existing knowledge-graph RAG framework developed by the German Aerospace Center (DLR). An ontology agent parses each query at runtime into a lightweight ontology of attribute-value pairs across five semantic dimensions: temporal scope, spatial scope, application domain, sensor/method type, and biophysical variable. A refinement agent then uses this ontology to filter the retrieved documents and information, producing a curated context. Finally, a generation agent drafts and refines the final answer. All agents run on the same Mistral Small 24B open-source model as the baseline.

We evaluated four pipeline variants against 70 EO questions: the two-step baseline, a zero-shot model, a no-ontology variant of our pipeline, and the full ontology-enhanced pipeline. Answers were individually rated by Claude Sonnet 4.6 on a 1--10 scale across eight criteria: scientific accuracy, relevance, completeness, directness, quantitative precision, response structure, scope awareness, and helpfulness. Our ontology-enhanced pipeline achieved an overall score of 7.48, ahead of the no-ontology variant (6.88), the zero-shot model (6.28), and the two-step baseline (5.51). It ranked first in all eight categories, with the largest gains over the no-ontology variant in helpfulness (+0.86), directness (+0.81), and scientific accuracy (+0.64). All pairwise comparisons were significant at $p<0.001$ with effect sizes greater than $r=0.77$. Quantitative precision remained the weakest criterion (5.93), largely because we only used document abstracts, which rarely contain exact numbers. The architecture can be applied to other fields and systems wherever queries can be described across multiple dimensions.