\chapter{Introduction}
\label{chap:introduction}

\section{Context and Motivation}
\label{sec:context}

Large language models (LLMs) hallucinate, and in scientific domains the consequences are more significant than in less critical settings. The most direct mitigation is to ground generation in retrieved evidence, the core idea behind Retrieval-Augmented Generation (RAG). A RAG system adds documents such as papers, articles, and web pages to the model's input, retrieved by term-based methods, semantic embeddings, or both. This approach has become central to reliable AI in scientific domains, where answers need correct terminology, quantitative precision, and explicit source attribution~\citep{lewis2020retrieval}.

Earth observation (EO) is a domain that is particularly sensitive to all of these. It spans remote sensing, climate monitoring, and environmental assessment, and it relies on specialised vocabulary, quantitative accuracy, and strict temporal and spatial context. Each measurement or study only makes sense in context: which sensors were used, what was measured, when and where, and under which conditions.

In this thesis, we investigate why larger proprietary language models such as GPT-5.1 and Claude Sonnet 4.5 handle EO queries well while open-source RAG implementations fall short. This is most evident in the two-step knowledge-graph RAG model developed by~\citet{elbaff2025earthobservationrag} at the German Aerospace Center (DLR). We use this model and the data generated by its architecture as the \emph{baseline} throughout this thesis. All proposed extensions are compared against it.

Standard RAG approaches, including the baseline, retrieve documents based on surface-level text similarity without modelling the semantic structure of the query. Even when relevant documents are retrieved, the assembled context fails to support generation, critical details are lost, and irrelevant passages dilute the answer to the point where it cannot support reliable scientific reasoning.

\section{Problem Statement}
\label{sec:problem}

After systematic manual analysis of baseline answers, the two-step baseline~\citep{elbaff2025earthobservationrag} proved to fail on specialised and complex queries due to a lack of deeper contextual awareness. The shortcomings identified below are expanded with detailed examples in Chapter~\ref{chap:preliminaries}.

\subsection{Key Shortcomings}
\label{subsec:shortcomings}

We evaluated how well the baseline model handled questions of varying complexity and type, focusing on domain-specific queries that combine multiple concepts and require both reasoning and specialised data. We compared the baseline against Anthropic's Claude Sonnet 4.5 and OpenAI's GPT-5.1. After an initial observation of the main faults, we refined the prompts and added extra instructions, before settling on four main shortcomings. Further examples and analysis are covered in Chapter~\ref{chap:preliminaries}.

\subparagraph{1. Delayed direct answers}

A direct question warrants a direct answer. Instead, the model opens with several sentences of background and context, only superficially related to the query. The proprietary models consistently opened with the answer itself, offering context only to support it.

\subparagraph{2. Missing quantitative precision}

The cited sources contain essential numbers and measurements, but the answers omit them. The model mentions relevant papers without extracting the quantitative details scientists need: thresholds, metrics, and the conditions under which those values were obtained.

\subparagraph{3. Lack of useful structure}

Not all questions are best answered with prose. The model favours long paragraphs where key points are scattered and mixed with background, regardless of whether a table or list would be more appropriate.

\subparagraph{4. Context misalignment}

Domain-specific questions combine multiple contexts and dimensions. The answer should highlight what is directly relevant and state clearly when sources address a different setting. Mixing contexts creates a risk of serious misunderstanding, especially when experts assume the reported information is specific to their query.

\section{Research Questions}
\label{sec:research-questions}

\textbf{Main Research Question:} Can a multi-agent architecture with dynamic ontology generation reduce hallucinations and improve answer quality in domain-specific RAG systems for Earth observation queries?

\textbf{Sub-Research Question 1:} How can query-specific ontologies capture the semantic structure of Earth observation questions to guide document retrieval and context refinement?

\textbf{Sub-Research Question 2:} To what extent can small language models handle complex Earth observation queries within a multi-agent RAG pipeline?

\textbf{Sub-Research Question 3:} How reliable are LLM-as-a-judge evaluation frameworks for assessing RAG systems in specialised scientific domains, and what are their key limitations?

\textbf{Sub-Research Question 4:} Does ontology-based semantic grounding provide significant performance benefits for small-model RAG systems, and to what extent can such an approach generalise to other knowledge-intensive domains beyond Earth observation?

The system we propose in Chapter~\ref{chap:approach} is designed to address all four.

\section{Thesis Contributions}
\label{sec:contribution}

This thesis contributes to open-source RAG systems for scientific domains in the following ways:

\begin{enumerate}
\item \textbf{Systematic diagnosis of baseline shortcomings.} Through structured manual analysis, we found four recurring failure modes in the two-step baseline: delayed directness, missing quantitative precision, weak response structure, and context misalignment.

\item \textbf{Design and implementation of an ontology-enhanced three-agent pipeline.} We designed a multi-agent architecture in which an ontology agent, a refinement agent, and a generation agent each enforce explicit semantic reasoning about the query, acting as guardrails against hallucination and context drift.

\item \textbf{Integration of query-specific ontologies with refinement and generation.} The pipeline generates a compact, query-specific ontology for each query rather than relying on a static domain taxonomy. This ontology covers five semantic dimensions: temporal scope, spatial scope, application domain, sensor/method, and biophysical variable, and is used to filter documents and guide answer construction.

\item \textbf{Tailored multi-criteria LLM-as-a-judge evaluation framework.} We developed an eight-criterion evaluation framework using Claude Sonnet~4.6 as judge on a 1--10 scale, addressing score saturation and context-length biases observed in earlier configurations. We compared four pipeline variants across 70~Earth observation questions.

\item \textbf{Per-criterion and qualitative error analysis.} The evaluation doesn't only include mean scores: per-criterion and per-question results are analysed alongside manual error investigation to identify both the strengths and remaining limitations of the ontology-enhanced approach.
\end{enumerate}

While developed for Earth observation, the principles of query-adaptive ontology generation and semantic document refinement apply to other domains where queries are inherently multi-dimensional and source context cannot be ignored.

\section{Thesis Structure}
\label{sec:structure}

The remainder of this thesis is organised as follows. Chapter~\ref{chap:background} covers background on RAG systems and Earth observation, and reviews related work on ontology-grounded retrieval, multi-agent systems, and LLM-based evaluation. Chapter~\ref{chap:preliminaries} deepens the analysis of the baseline model's shortcomings with detailed examples. Chapter~\ref{chap:approach} presents the three-agent architecture, specifying each agent's role and showing how the design addresses each identified shortcoming. Chapter~\ref{chap:evaluation} describes the evaluation framework: the dataset of 70~EO questions, four pipeline variants, and the multi-criteria LLM-as-a-judge methodology. Chapter~\ref{chap:results} reports quantitative results, compares the four variants, and identifies failure modes through detailed examples. Chapter~\ref{chap:discussion} interprets the results and discusses limitations. Finally, Chapter~\ref{chap:conclusion} reflects on possible future work and concludes by revisiting the research questions.
